### 2018-05-09 Final Paper/paper.tex
Auto stash before merge of "master" and "origin/master"

	% As a general rule, do not put math, special symbols or citations
	% in the abstract
	\begin{abstract}

@@

\section{Introduction}

\section{The Need for Smoothing}

%\begin{figure}[!h]
%	\centering
%	\includegraphics[width=\linewidth]{{"plots/unsmoothed_data"}.png}
%	\caption{Look at this mess. Look upon it, and weep.}
%	\label{fig:unsmoothed_data}
%\end{figure}

\section{Windowing}
One of the core problems we needed to solve before tackling any others was making our data smoother. Initially, all of the data we worked with was extremely noisy, and yielded problems when we initially attempted to detect peaks or find accurate periods. In pursuing this, all other computation related to the project run would run much more efficiently and with less error. The primary method that we started smoothing with filters. The first that we got to work properly was the Savitzky-Golay filter, one developed by increasing the signal-to-noise ratio of the plot without entirely corrupting the signal. To do this it essentially generates new fitted subsets of the data according to an equation. While this method has been working pretty effectively for us so far, we wanted to find out whether or not we could find better results by trying to create new filters based on different functions. For this, we turned to window functions; methods that are designed to fit data zero-valued/normalized according to a given interval, using an equation. These were normally used in signal processing, but, could potentially be used to smooth any plot that was given as a signal, depending on its purpose.

However, most of these proved ineffective. The windowing methods we tested were primarily the ones most commonly used in signal processing/general usage. These included, the Hamming window function, the Blackman Harris, the Flat-Top, and the Bartlett-Hann functions. However, all of them yielded problematic results, as they would not give us accurate period detection, no matter what we tried. Considering the nature of these functions (with most of them being derivatives of a core few, which we tested for), we then concluded that these would be ineffective for our purposes.

We had also tried to apply a few methods of dealing with outliers in the data, however, actual filters trumped those as well.

\section{Genetic Algorithms}

One of the most common ways to tackle the period-finding problem is to do curve fitting. That is, find the correct parameters to a function in order to get the difference between that function’s outputs and the true data as small as possible. Then, we extract the period from that function (since it should be easy to mathematically determine the period from a set function). Although this method works extremely well, it should also be clear that finding the correct parameters for the function is prohibitively time-consuming (since the potential options are infinite, and testing each set of parameters requires an evaluation of the function at every time step). However, given how common this approach is (and how common the parameter-finding problem is), a large number of algorithms exist to get approximate best parameters in more reasonable computation times.
For this problem, we chose to use a genetic algorithm, which is one of the most common ways to do parameter-finding (also, Mathias had working code from Systems Biology I, so it was relatively easy to implement). Genetic algorithms are meant to emulate the way that evolution works (in the loosest sense of the word). To do this, they take a random sampling of the parameter space, to get the parameters for the first group of parents (each of which is simply a full parameter set for the function, so for example, for Equation \ref{eq:mFourFit}, we would have a C, five As, five {$\phi$}s, and a $\tau$). The next step is to evaluate the quality of each parameter set using a cost function (for us, this was simply a least-squares difference between our data and the function defined with each parameter set), and sort the parents according to the quality of their fit. Once the parents have been sorted, we take a subset of the parents according to a selection strategy. What we used for this project was tournament selection - for each parent that we pick, we first pick a random subset of the parents, then we take the best parent from that subset. For this version of a genetic algorithm, we pick the same number of parents as we had in the last round - we just pick the best parents multiple times, and exclude the worst parents, by design. This whole process is then repeated for either a set number of generations (in our case, 10-20), or until the error for the best parameter set becomes lower than some threshold. The best parameter set at the end of this process then becomes the parameters that we use for the function.

@@

\section{Peak Finding with Derivatives}

\section{Our Peak Detection Process}

@@

\section{Conclusion}

>>>>

	% As a general rule, do not put math, special symbols or citations
	% in the abstract
	\begin{abstract}
		For this project, we were tasked with analyzing oscillations of Per2 and BMal1 in data recorded from cells damaged by cancer - specifically, we wanted to find the period of the oscillations. Michelle Farkas, from the University of Massachusetts Amherst Department of Chemistry, provided us data from cancer-damaged cells. This data was highly irregular and noisy, making a straightforward computational analysis impossible. However, by using a hand-made suite of data cleaning tools and a combination of broad peak searching and sensitive peak detection, we were able to create a piece of software that consistently finds all of the peaks and troughs in the data that was provided to us. This allowed us to produce a function of the period over time for individual datasets.

@@

\section{Introduction}
The study of circadian clocks within organisms has been a growing field in systems biology.  We specifically looked at the oscillations of Per2 and BMal1 proteins in cells damaged by cancer.  Michelle Farkas, from the University of Massachusetts Amherst Department of Chemistry, provided us data from cancer-damaged cells that were starved for 24 hours, 12 hours, and not starved at all.  It's important to understand the period of these oscillations as knowing the period plays into understanding the mechanics of the lager celular systems.  However, due to the nature of the data there were no consistently accurate automated peak detection system in place.  The data has a couple of idiosyncrasies that made automated analysis harder - the oscillations are damped, there was a downwards trend in all of the datasets, and there was a lot of noise in the data. Furthermore, most period finding algorithms are designed to work for data that has a constant period, but we found that this data had a period that varied with time.
Moving forward with the task of determining the period we proceeded with several initial steps.  This included analysing smoothing techniques as we needed to smooth the data while ensuring that it remained as true to the actual data as possible through manipulations and derivatives.   Additionally, the use of genetic algorithms to create a multiple cosines function for obtaining the periodicity was explored.  The classic use of derivatives for finding peaks to calculate the period was also investigated for determining the best automated period detection tool.

\section{The Need for Smoothing}
\begin{figure}[!h]
	\centering
	\includegraphics[width=\linewidth]{{"plots/unsmoothed_data"}.png}
	\caption{Look at this mess. Look upon it, and weep.}
	\label{fig:unsmoothed_data}
\end{figure}

\section{Windowing}
One of the core problems we needed to solve before tackling any others was making our data smoother. Initially, all of the data we worked with was extremely noisy, and yielded problems when we initially attempted to detect peaks or find accurate periods. In pursuing this, all other computation related to the project run would run much more efficiently and with less error. The primary method that we started smoothing with filters. The first that we got to work properly was the Savitzky-Golay filter, one developed by increasing the signal-to-noise ratio of the plot without entirely corrupting the signal. To do this it essentially generates new fitted subsets of the data according to an equation. While this method has been working pretty effectively for us so far, we wanted to find out whether or not we could find better results by trying to create new filters based on different functions. For this, we turned to window functions; methods that are designed to fit data zero-valued/normalized according to a given interval, using an equation. These were normally used in signal processing as these algorithms are usually geared towards frequency responses and plots for proper spectral analysis (primarily detecting things like spectral leakage and other frequency/sinusoid related plots and problems). Yet, these algorithms could potentially be used to smooth any plot that was given as a signal, depending on its purpose.

However, most of these proved ineffective. The windowing methods we tested were primarily the ones most commonly used in signal processing/general usage. These included the Hamming window function, the Blackman Harris, the Flat-Top, and the Bartlett-Hann functions.

\begin{figure}[!h]
	\centering
	\includegraphics[width=\linewidth]{{"plots/grid"}.jpg}
	\caption{Hamming and Flattop Window Test: Using MATLAB code, this shows us one of the sets of data, and shows the effect of using the Hamming and Flattop window functions to dampen it. The bottom result is supposed to show us the periods detected after the application of the algorithm, however, it is not detecting the periods properly, likely due to the cancerous data. This is further confirmed by the period being detected, which is much larger than it should be, as the periods should be 20-30h at best. The explicit use versus the ones matrix indicates if there was an additional matrix used in mitigating the effects of the algorithm on the signal. Explicit use meant no additional matrix was used, ones matrix indicates that ones matrix was used.}
	\label{fig:grid}
\end{figure}

However, all of them yielded problematic results, as they would not give us accurate period detection, no matter what we tried. Considering the nature of these functions (with most of them being derivatives of a core few, which we tested for), we then concluded that these would be ineffective for our purposes, at least in the timeframe we were working in.

We had also tried to apply a few methods of dealing with outliers in the data, previously working with two methods for removing outliers based on standard deviation, and then a custom spike smoothing method which is intended to isolate where data ‘spikes’ and then modify them according to the adjacent points in the plot, rather than an outright removal. This would have us then decide whether or not smoothing or removal is better for our data in the preliminary stages of our analyses. However, actual filters trumped those as well, but, this was not totally unexpected as the filters we were previously working with were industry standard for smoothing.

\section{Genetic Algorithms}
One of the most common ways to tackle the period-finding problem is to do curve fitting. That is, find the correct parameters to a function in order to get the difference between that function’s outputs and the true data as small as possible. Then, we extract the period from that function (since it should be easy to mathematically determine the period from a set function). Although this method works extremely well, it should also be clear that finding the correct parameters for the function is prohibitively time-consuming (since the potential options are infinite, and testing each set of parameters requires an evaluation of the function at every time step). However, given how common this approach is (and how common the parameter-finding problem is), a large number of algorithms exist to get approximate best parameters in more reasonable computation times.
For this problem, we chose to use a genetic algorithm, which is one of the most common ways to do parameter-finding (also, Mathias had working code from Systems Biology I, so it was relatively easy to implement). Genetic algorithms are meant to emulate the way that evolution works (in the loosest sense of the word). To do this, they take a random sampling of the parameter space, to get the parameters for the first group of parents (each of which is simply a full parameter set for the function, so for example, for Equation \ref{eq:mFourFit}, we would have a C, five As, five {$\phi$}s, and a $\tau$). The next step is to evaluate the quality of each parameter set using a cost function (for us, this was simply a least-squares difference between our data and the function defined with each parameter set), and sort the parents according to the quality of their fit. Once the parents have been sorted, we take a subset of the parents according to a selection strategy. What we used for this project was tournament selection - for each parent that we pick, we first pick a random subset of the parents, then we take the best parent from that subset. For this version of a genetic algorithm, we pick the same number of parents as we had in the last round - we just pick the best parents multiple times, and exclude the worst parents, by design. This whole process is then repeated for either a set number of generations (in our case, 10-20), or until the error for the best parameter set becomes lower than some threshold. The best parameter set at the end of this process then becomes the parameters that we use for the function.

@@

\section{Peak Finding with Derivatives}
Standard peak finding techniques have traditionally relied on the first derivative to find peaks and valleys.  This process looks for where the first derivative crosses zero and depending on the direction a peak (positive to negative) or valley (negative to positive) is found.  When applying this technique the first derivative correctly detects most peaks and valleys.  However, due to damping and the downward trend of the data it fails to find peaks that optically appear to be significant in the systems period.  This can be seen in Figure \ref{fig:first_drv_fail} as in between the first and last valley peak pairing there appears to be an oscillation in the data that doesn't fully pass the first derivatives threshold.  Thus to find the optically noticeable peaks we needed a more sensitive measure for which we turned to the third derivative.  The Third derivative operates in the same manner as the first derivative for finding peaks and valley but it also looks at the curvature of the data as well.  The third derivative correctly detects peaks and valley including the visually significant peaks and valleys.  However, it also detects peaks and valleys where there don't appear to be any significant to the periodicity which can be seen in Figure \ref{fig:third_drv_fail}.  This signified a hybrid method with a sensitivity greater than the pure first derivative and less than the pure third derivative would be required.

\begin{figure}[!h]
	\centering
	\includegraphics[width=\linewidth]{{"plots/first_drv_fail"}.png}
	\caption{Using the first derivative to find peaks.}
	\label{fig:first_drv_fail}
\end{figure}

\begin{figure}[!h]
	\centering
	\includegraphics[width=\linewidth]{{"plots/third_drv_fail"}.png}
	\caption{Using the third derivative to find peaks.}
	\label{fig:third_drv_fail}
\end{figure}

\section{Our Peak Detection Process}
The final peak detection method utilized correcting data through polynomial fitting, the first derivative, and the third derivative.  First the data was smoothed using the Savitzky-Golay filter to remove noise from the data.  A third degree polynomial was fit to the dataset, thus identifying the major damping trend.  This third degree polynomial was then inverted and added to the actual data thereby removing the overarching damping and creating a corrected dataset.  By removing the overarching damping trend optical peaks became more susceptible to detection by the first derivative. The peak and valley identification subsequently began by using the first derivative to identify peaks and valleys in the corrected dataset.  The average time between peaks and valleys was then calculated and if a gap between peaks and valleys larger than 1.33 times the average distance was detected it was marked as missing a peaks and/or valleys.  The third derivative was then applied to the uncorrected data strictly within identified gaps to find the remaining peaks and valleys.  The peaks and valleys initially found with the first derivative of the corrected data and those found using the third derivative on the uncorrected data in identified gaps were then consolidated on the uncorrected data to return the set of final peaks and valleys.  With the final set of peaks and valleys the period could then be calculated.  This process is shown in Figure \ref{fig:pv_det_final}.

\begin{figure}[!h]
	\centering
	\includegraphics[width=\linewidth]{{"plots/pv_det_final"}.png}
	\caption{Our final peak and valley detection process. This process worked consistently for every dataset that we tried it on.}
	\label{fig:pv_det_final}
\end{figure}

@@

\section{Conclusion}
By using a hand-made suite of data cleaning tools and a combination of broad peak searching and sensitive peak detection, we were able to create a piece of software that consistently finds all of the peaks and troughs in the data that was provided to us. We ended up determining that finding the period in data like this is extremely difficult, particularly because noise makes the derivatives useless, the downward trend of the graph makes some peaks not show up in derivative analysis, and the damping of the oscillations over time makes the last part of the data very hard to analyze. We also noticed that the fact that the period changes over time makes the coding problem harder, because we cannot simply assume that the peaks will be consistently spaced throughout the data, so placing constraints can end up being harder.
However, by smoothing the data using the Savitzky-Golay filter, we removed our issues with noise entirely. We tried using a Genetic Algorithm to fit a combination of multiple cosines to the data, using both the mFourFit (Equation \ref{eq:mFourFit}) and FFT-NLLS (Equation \ref{eq:FFT-NLLS}) equations. However, this simply provided us with a general trend fitting, and not a oscillation fitting. We were able to use polynomial fitting to get the same general trend fitting as the Genetic Algorithms, but with a much, much faster fitting time (10 hours compared to <1 second). Using that trend fit, we corrected the downwards slope of the data. We then used first-derivative peak detection to cast a wide net and get the most obvious peaks. By using the mean of the time between peaks, we were able to find places where there should be missing peaks (anywhere the time between found peaks was bigger than $1.3$ times the mean). Inside of those areas, we then used the much more sensitive third-derivative peak detection, which consistently found all of the missing peaks and troughs that first-derivative didn’t find. Having a consistent way to find the peaks and troughs in the data allowed us to plot the period over time for all of the data sets that we were given.
